---
title: "Project 1"
author: "Anastasiia Gmyrina"
editor: visual
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    df-print: paged
    smooth-scroll: true
    embed-resources: true
    grid:
      body-width: 800px
---

## Approach

For this project, I will work with the chess tournament text file and turn it into a clean dataset. The main challenge will be working with a text file where the information is spread across multiple lines. There are a few patterns in the file. For example, each player’s information is spread across two lines followed by a separator, and the round results contain information about the player’s opponents. I will use these patterns to extract the required information, create a dataframe, and export the final results as a CSV file.

```{mermaid}
flowchart LR
    A(Read TXT file) --> B(Clean and structure data)
    B --> C(Extract player and opponent information)
    C --> D(Calculate average opponent rating)
    D --> E(Create dataframe and export CSV)
```

## Read txt file

Since each player's information is spread across multiple lines, I used `readLines()` to read file line by line.

```{r}
#| message: false
#| warning: false
library(tidyverse)
chess <- readLines("tournamentinfo.txt")
head(chess,10)
```

I removed the first four lines since it's just the headers, and checked the first 10 lines again to make sure the dataset now starts with the first player record.

```{r}
chess_data <- chess[-(1:4)]
head(chess_data, 10)
```

## Clean and structure data

Since `readLines()` stores each line of the text file as an element in a vector, we can use vector indexing to select specific lines by their position. We found a repeating pattern where the player information is on lines 1, 4, 7, etc., and the additional details are on lines 2, 5, 8, etc. We can use `seq()` to select these lines and store them in two separate vectors.

```{r}
player_info <- chess_data[seq(1, length(chess_data), by = 3)]
head(player_info,3)
```

```{r}
details_info <- chess_data[seq(2, length(chess_data), by = 3)]
head(details_info,3)
```

Since player_info is currently a vector, I converted it into a tibble so I can use separate_wider_delim() to split the player information into separate columns.

```{r}
players <- tibble(player_info = player_info)
players <- players %>%
  separate_wider_delim(player_info,delim = "|", names = c("player_num", "player_name", "points","round1", "round2", "round3", "round4", "round5", "round6", "round7"),too_many = "drop")
head(players)
```

```{r}
details <- tibble(details_info = details_info)
details <- details %>%
  separate_wider_delim(details_info, delim = "|", names = c("state", "rating_info"),too_many = "drop")
head(details)
```

The rating_info column contains the player ID, pre-rating, and post-rating in one string. I will use str_extract() to get the pre-rating, convert it to numeric and keep only two columns I need.

```{r}
details <- details %>% mutate(pre_rating = as.numeric(str_extract(rating_info, "(?<=R: )\\d+"))) %>% select(state, pre_rating)

head(details)
```

## Extract player and opponent info

The round columns contain the game result and the opponent's player number. I used pivot_longer() to combine the seven round columns into rows, so each round can be analyzed separately.

```{r}
opponents <- players %>%
  pivot_longer(
    cols = round1:round7,
    names_to = "round",
    values_to = "result")

opponents <- opponents %>%
  mutate(opponent = str_extract(result, "\\d+"))
head(opponents, 10)
```

## Calculate average opponent rating

Since the rows in `details` are in the same order as the player numbers, I used row_number() to add the corresponding player number to each row. This allows each player's pre-rating to be matched with their player number.

```{r}
details <- details %>% mutate(player_num = row_number())
head(details,2)
```

Now opponents has the opponent number, and `details` has the player number and pre-rating. I need to match them.

```{r}
opponents <- opponents %>%
  mutate(opponent = as.numeric(opponent)) %>%
  left_join(details, by = c("opponent" = "player_num"))
head(opponents,3)
```

Now we can calculate the average opponents rating for each of the players

```{r}
avg_ratings <- opponents %>%
  group_by(player_num) %>%
  summarise(avg_opponent_rating = round(mean(pre_rating, na.rm = TRUE)))
head(avg_ratings,3)
```

## Create a dataframe and export to CSV

I combined the player information with the average opponent ratings and player details using LEFT_JOIN

```{r}
players <- players %>%
  select(player_num, player_name, points) %>%
  left_join(avg_ratings, by = "player_num") %>%
  mutate(player_num = as.numeric(player_num)) %>%
  left_join(details, by = "player_num")

head(players, 3)
```

Finally, I selected and organized the columns to match the assignment requirements and exported it as csv file.

```{r}
players <- players %>%
  select(player_name, state, points, pre_rating, avg_opponent_rating)

head(players, 3)
write_csv(players, "chess_players.csv")
```
