---
title: "NYC crashes and SHAP analysis"
author: "Anastasiia Gmyrina"
format: html
editor: visual
---

## Introduction

This analysis uses NYC motor vehicle collision data to demonstrate how SHAP can help explain a machine learning model. I use an XGBoost model to predict whether a crash resulted in an injury and then use SHAP to examine which features influenced the model's predictions.

## 1. Load Libraries and Data

```{r}
#| message: false
#| warning: false
library(xgboost)
library(shapviz)
library(tidyverse)
url <- paste0(
  "https://data.cityofnewyork.us/resource/h9gi-nx95.csv?",
  "$limit=10000"
)

library(xgboost)
library(shapviz)
library(tidyverse)

crashes <- read_csv("nyc_crashes_10000.csv")
```

The dataset contains 10,000 motor vehicle collision records from NYC Open Data.

## 2. Create the Injury Variable

I created a binary variable called `injury`. A value of 1 means that at least one person was injured in the crash, while 0 means that no injuries were reported.

```{r}
crashes <- crashes %>%
  mutate(
    injury = if_else(number_of_persons_injured > 0, 1, 0)
  )

crashes %>%
  count(injury)

```

## 3. Prepare Variables for the Model

For this demonstration, I selected four predictors: borough, hour of the crash, vehicle type, and the primary contributing factor.

```{r}
model_data <- crashes %>%
  transmute(
    injury = factor(injury),
    borough = factor(borough),
    hour = as.integer(substr(crash_time, 1, 2)),
    vehicle_type = factor(vehicle_type_code1),
    contributing_factor = factor(contributing_factor_vehicle_1)
  ) %>%
  drop_na()
model_data %>%
  glimpse()
```

## 4. Split the Data

I used 80% of the observations for training and 20% for testing.

```{r}
set.seed(123)

split <- sample(
  seq_len(nrow(model_data)),
  size = 0.8 * nrow(model_data)
)

train_data <- model_data[split, ]
test_data <- model_data[-split, ]
```

## 5. Prepare Data for XGBoost

The categorical variables need to be converted into numeric dummy variables before they can be used by XGBoost.

```{r}
x_train <- model.matrix(injury ~ . - 1, data = train_data)
x_test <- model.matrix(injury ~ . - 1, data = test_data)

y_train <- as.numeric(train_data$injury) - 1
y_test <- as.numeric(test_data$injury) - 1

```

## 6. Train the XGBoost Model

I trained an XGBoost model using the training data and generated predictions for the test data.

```{r}
model <- xgboost(
  x = x_train,
  y = factor(y_train),
  nrounds = 100
)

predictions <- predict(model, x_test)

head(predictions)
```

## 7. Evaluate the Model

Predicted probabilities above 0.5 were classified as injury predictions.

```{r}
results <- tibble(
  actual = y_test,
  predicted = if_else(predictions > 0.5, 1, 0)
)

results %>%
  summarise(
    accuracy = mean(actual == predicted)
  )


```

The model achieved approximately **68% accuracy** on the test data. The purpose of this analysis is not to optimize prediction accuracy, but to demonstrate how SHAP can be used to interpret a machine learning model.

## 8. Calculate SHAP Values

SHAP values were calculated for the observations in the test dataset.

```{r}
shap_values <- predict(
  model,
  x_test,
  type = "contrib"
)

shap_matrix <- shap_values[, -ncol(shap_values)]
colnames(shap_matrix) <- colnames(x_test)

shap <- shapviz(
  shap_matrix,
  X = x_test
)

```

## 9. Overall Feature Importance

The SHAP importance plot shows which features had the largest influence on the model's predictions overall

```{r}
sv_importance(shap)
```

In this model, **hour of the crash** was the most important feature overall. Several contributing factors, borough indicators, and vehicle types were also important.

## 10. Explain One High-Risk Prediction

I selected the observation with the highest predicted probability of injury.

```{r}
high_risk <- which.max(predictions)

predictions[high_risk]

```

The SHAP waterfall plot shows how the individual features contributed to this prediction.

```{r}
sv_waterfall(
  shap,
  row_id = high_risk
)
```

Finally, I checked the original test observation and its actual outcome.

```{r}
test_data %>%
  slice(high_risk)

y_test[high_risk]
```

For this observation, the model predicted a very high probability of injury, and the actual outcome was also an injury. The waterfall plot shows that **bike involvement** and **the hour of the crash** were two of the strongest positive contributors to this prediction.

## Summary

This example demonstrates the difference between prediction and explanation. XGBoost produced the injury prediction, while SHAP helped show which features the model relied on when making that prediction.

SHAP explains **model behavior**, not causation. A high SHAP value for a feature does not mean that the feature caused the injury; it means that the feature influenced the model's prediction.

## Sources

NYC Open Data. *Motor Vehicle Collisions – Crashes*. This dataset contains NYPD-reported motor vehicle collisions in New York City and was the source of the crash data used in this analysis.

[NYC Open Data — Motor Vehicle Collisions](https://data.cityofnewyork.us/d/h9gi-nx95?utm_source=chatgpt.com)

Lundberg, S. M., & Lee, S.-I. (2017). *A Unified Approach to Interpreting Model Predictions*. This paper introduced SHAP as a method for explaining individual machine learning predictions.

[SHAP paper — Lundberg & Lee (2017)](https://papers.neurips.cc/paper/7062-a-unified-approach-to-interpreting-model-predictions?utm_source=chatgpt.com)

XGBoost documentation. XGBoost was used to build the classification model that predicted whether a crash resulted in an injury.

[XGBoost Documentation](https://xgboost.readthedocs.io/?utm_source=chatgpt.com)

The `shapviz` R package was used to visualize the SHAP values, including feature importance, dependence, and waterfall plots.

[shapviz R Package](https://cran.r-project.org/package=shapviz?utm_source=chatgpt.com)

### AI Assistance

ChatGPT (GPT-5.6 Sol) was used to help with R implementation, troubleshooting the XGBoost and SHAP workflow, and interpreting the SHAP output.
