---
title: "Assignment 2a"
format: html
editor: visual
---

# Approach

I selected six 2025 movies nominated for Best picture at the recent Oscars and collected ratings from five people using a 1–5 scale. Not every participant had seen every movie, so some ratings are missing, i will treat them as unavailable responses.

I will store data in SQLite database in three tables: users, movies, and ratings. Then I will use a JOIN to combine the tables into a nicer format with names, movie titles, and rating. I plan to load SQL data into R as a dataframe and then summarize the ratings and calculate statistical summary for each movie.

# SQL database

I first created movie_ratings.db in SQLite with three relational tables: users, movies and ratings. I used separate tables to avoid repeating user and movie information and connected them using their ids. I then used `INSERT INTO` to populate each table with the collected data and a `JOIN` to combine all three tables into a readable dataset.

```{sql}
--| eval: false
CREATE TABLE movies (
    movie_id INTEGER PRIMARY KEY,
    title TEXT NOT NULL
);

CREATE TABLE users (
    user_id INTEGER PRIMARY KEY,
    name TEXT NOT NULL
);

CREATE TABLE ratings (
    user_id INTEGER,
    movie_id INTEGER,
    rating INTEGER,
    FOREIGN KEY (user_id) REFERENCES users(user_id),
    FOREIGN KEY (movie_id) REFERENCES movies(movie_id)
);

INSERT INTO movies (movie_id, title)
VALUES
    (1, 'Sinners'),
    (2, 'One Battle After Another'),
    (3, 'Frankenstein'),
    (4, 'Hamnet'),
    (5, 'Marty Supreme'),
    (6, 'F1');

INSERT INTO users (user_id, name)
VALUES
    (1, 'Ana'),
    (2, 'Joe'),
    (3, 'Christine'),
    (4, 'Anna'),
    (5, 'Aika');
INSERT INTO ratings (user_id, movie_id, rating)
VALUES
    (1, 1, 5),
    (1, 2, 5),
    (1, 3, NULL),
    (1, 4, 4),
    (1, 5, 4),
    (1, 6, NULL),

    (2, 1, 3),
    (2, 2, 4),
    (2, 3, 5),
    (2, 4, 4),
    (2, 5, 3),
    (2, 6, 3),

    (3, 1, 3),
    (3, 2, 3),
    (3, 3, NULL),
    (3, 4, 3),
    (3, 5, 4),
    (3, 6, 5),

    (4, 1, 5),
    (4, 2, NULL),
    (4, 3, 5),
    (4, 4, 4),
    (4, 5, 5),
    (4, 6, 5),

    (5, 1, 4),
    (5, 2, 3),
    (5, 3, 4),
    (5, 4, NULL),
    (5, 5, 4),
    (5, 6, 2);
SELECT
    users.name,
    movies.title,
    ratings.rating
FROM ratings
JOIN users
    ON ratings.user_id = users.user_id
JOIN movies
    ON ratings.movie_id = movies.movie_id;

```

I connected R directly to my SQLite database using DBI and RSQLite. I used dbGetQuery() to run a SQL JOIN and load the combined results into R as a dataframe called `movie_ratings`

```{r}
#| message: false
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
```

```{r}
mydb <- dbConnect(SQLite(), "movie_ratings.db")
```

```{r}
movie_ratings <- dbGetQuery(mydb, "
SELECT
    users.name,
    movies.title,
    ratings.rating
FROM ratings
JOIN users
    ON ratings.user_id = users.user_id
JOIN movies
    ON ratings.movie_id = movies.movie_id
ORDER BY users.user_id, movies.movie_id;
")
glimpse(movie_ratings)

```

## Missing values

Not every participant had seen every movie, so some ratings are missing. Missing ratings from SQL were imported into R as NAs.

```{r}
movie_ratings %>%
  filter(is.na(rating))
```

## Movie summary

```{r}
movie_summary <- movie_ratings %>%
  group_by(title) %>%
  summarise(
    n_ratings = sum(!is.na(rating)),
    mean_rating = round(mean(rating, na.rm = TRUE), 2),
    median_rating = round(median(rating, na.rm = TRUE), 2),
    sd_rating = round(sd(rating, na.rm = TRUE), 2)
  )

movie_summary

```

I also created a bar chart showing the average rating per movie, ordered from highest to lowest.

```{r}
ggplot(movie_summary, aes(x = reorder(title, mean_rating), y = mean_rating)) +
  geom_col() +
  coord_flip()
```

# User summary

```{r}
user_summary <- movie_ratings %>%
  group_by(name) %>%
  summarise(
    n_ratings = sum(!is.na(rating)),
    mean_rating = round(mean(rating, na.rm = TRUE), 2)
  )

user_summary
```

# Results

The results show that the movies received different numbers of ratings because not every participant had seen every movie. Missing ratings were kept as NA and excluded from the statistical calculations. The summary tables make it easy to compare movie-level and user-level rating patterns.

```{r}
dbDisconnect(mydb)
```
